Real Time Variational Density Forecasts

### 2017-02-19 Confirmation/confirmation_V2.tex
tex updates

Electricity prices have been subject to an extensive forecasting literature and is known to be extremely volatile. For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of oberservations less than \$100/mWh, but on 39 occaisions the price spiked greater than \$1,000/mWh, reaching as high as \$9,974/mWh on November 29, 2012. For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period is more important than the mean value. To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it can average over parameter and model uncertainty that is often ignored in frequentist methods. See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for density forecasting in general.

Bayesian methods are computationally complex as they require integration over the set of unknown parameters, which must be large in a model that can capture the strong seasonal patterns in short term electricity load \citep{Taylor2003}, which is the main driver for electricity price and the focus of this work. To avoid this problem there are a range of techniques available to obtain the predictive distribution which do not involve integration, notably Markov Chain Monte Carlo (MCMC) methods. While MCMC has very strong convergence properties, computation is slow for even moderately complex models, which is problematic as Electricity demand in Victoria operates on a five minute interval. The predictive density for the next period's electricity load may not be available before it is observed.

Alternatives to MCMC include Variational Bayes (VB), which aim to replace the unknown true posterior distribution for a vector of unknown parameters $\theta$ with a parametric approximation. Variational Bayes will be explored as an alternative computation strategy to MCMC in the context of short term electricity load forecasting using data from the 2011-2015 period in Victoria. The advantages in computation time for VB will be compared to the loss of statisitcal accuracy associated with the use of an approximation of the posterior instead of the truth.

@@

Exact Bayesian inference in the context of forecasting uses the predictive density of $Y_{T+1}$ conditioned on the observed series, $y_{t}, t = 1, \dots T$, which is denoted by $y_{1:T}$. The first step requires the posterior distribution of the model parameters $\theta$, $p(\theta | y)$, where

\begin{equation}
\label{posterior}
 p(\theta | y_{1:T}) = \frac{p(y_{1:T}|\theta)p(\theta)}{\int_\theta p(y_{1:T}|\theta)p(\theta) d\theta}.
\end{equation}

Generally the solution to (\ref{posterior}) is unavailable. This section will offer two alternatives, MCMC and VB.
MCMC is used to create a sample from $p(\theta | y)$, and any function of interest from $p(\theta | y)$ can be found with the relevant function applied to the sample, while VB replaces $p(\theta | y)$ with an approximation $q(\theta | \lambda)$, where $\lambda$ is a vector of auxillary parameters associated with the approximation and not the model. For example, if $q(\theta | \lambda)$ is Gaussian $\lambda$ would be the mean and variance of $\theta$. Variational Bayes aims to choose the distribution $q(\theta | \lambda)$ such that the Kullback-Leibler divergence from $q(\theta | \lambda)$ to $p(\theta | y)$ is minimised.

@@

There are many types of MCMC algorithm, the most common being a Gibbs sampler which iteratively samples from a $k$ dimensional parameter vector $\theta$ via the conditional distributions

@@

With enough iterations, these samples converge in distribution to $p(\theta | y)$ and can be used for inference. The computation time per iteration and number of iterations required for convergence is problem specific but typically increases with model complexity.

@@

\begin{equation}
\label{AR2}
y_t = \rho_1 y_{t-1} + \rho_2 y_{t-2} + \epsilon_t
\end{equation}

where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. The likelihood of parameters $\theta = (\rho_1, \rho_2m \sigma^2)'$ is given by

\begin{align}
\label{likelihood}
L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( \textbf{y}' \Sigma^{-1} \textbf{y} + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}
\end{align}

@@

\begin{equation}
\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1), \gamma(0) \end{array} \right],\
\end{equation}

@@

\begin{align}
p(\sigma^2) &\propto \sigma^{-2} \nonumber \\
p(\rho_1, \rho_2) &\propto \mathbb{I}(\rho_2 > -1)\mathbb{I}(\rho_2 < 1 + \rho_1) \mathbb{I}(\rho_2 < 1 - \rho_1) \nonumber
\end{align}

are used, where $\mathbb{I}$ is the indicator and $p(\rho_1, \rho_2)$ defines the AR(2) stationary region.

The conditional distribution $p(\sigma^2 | y, \phi_1, \phi_2)$ is Inverse Gamma, while the conditionals for both $\phi$ parameters are intractable and sampled using a Metropolis-Hastings step with truncated normal random walk proposals. The truncation ensures that draws are only taken from the AR(2) stationary region.

@@

A faster alternative to MCMC is Variational Bayes, which introduces an approximating distribution $q(\theta | \lambda)$ choosing the functional form $q$ and given auxillary parameters $\lambda$ so that  misimises the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\theta | \lambda)$ the true posterior $p(\theta | y)$. The KL divergence is defined by

\begin{equation}
\label{KL-def}
KL[q(\theta | \lambda)||p(\theta | y)] = \int q(\theta | \lambda) \ln \left( \frac{q(\theta | \lambda)}{p(\theta | y)}\right) d\theta.
\end{equation}

The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\theta | y)$ and $q(\theta | \lambda)$ that will equal zero if and only if $p(\theta | y) = q(\theta | \lambda)$ almost everywhere \citep{Bishop2006}. Note that $KL[q(\theta | \lambda)||p(\theta | y)]$ can be expressed as

\begin{equation}
\label{KL-ELBO}
KL[q(\theta | \lambda)||p(\theta | y)] = \ln(p(y)) - \mathcal{L}(q, y)
\end{equation}

where $\mathcal{L}(q, y)$ is known as the Evidence Lower Bound (ELBO), defined by

\begin{equation}
\label{ELBO}
\mathcal{L}(q, y) = \int_{\theta} q(\theta|\lambda) \ln (p(y, \theta|\lambda)) d\theta -  \int_{\theta} q(\theta|\lambda) \ln (q(\theta|\lambda)) d\theta.
\end{equation}

Since $\ln(p(y))$ is constant given $y$, maximising (\ref{KL-ELBO}) with respect to $q$ is equivalent to minimising (\ref{KL-def}). Maximising the ELBO is much more convenient than minimising the KL Divergence, which can be seen in the following sections based on the two major implementations of Varitaional Bayes: Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these take advantage of the functional form of the ELBO and offer computationally efficient algorithms to find a $q(\theta | \lambda)$ that is optimal within a particular clas of distributions.

@@

\begin{equation}
\label{mf1}
q(\theta|\lambda) = \prod_i^k q_i(\theta_i | \lambda_i).
\end{equation}

Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars. Each $\theta_i$ has an associated vector $\lambda_i$ which may be of a higher dimension than $\theta_i$. $\lambda_i$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\lambda_i)$.

MFVB is widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}), particularly in exponential family models \citep{Wainwright2008}.  Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the computation is simple. MFVB expresses (\ref{ELBO}) as a function of one $q_i$ only and then maximises with respect to that $q_i$. The final algorithm requires this to be repeated for each $i$ in $1, \dots, k$. Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that

\begin{equation}
\label{mf2}
q(\theta_i | \lambda_i) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])
\end{equation}

For illustration, consider data $y_i$ for $i = 1, \dots, n$ generated from a $\mathcal{N}(\mu, \sigma^2)$ distribution with priors $p(\mu) \sim \mathcal{N}(\gamma, \lambda)$ and $p(\sigma^2) \sim Inv.Gamma(\alpha, \beta)$. It can be shown that

\begin{equation}
\label{mf3}
q(\mu) \propto \exp \left\{ \frac{-(n\lambda + \mathbb{E}(\sigma^{-2}))}{2\lambda\mathbb{E}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}(\sigma^{-2})\gamma + \lambda \sum_{i=1}^{n} y_i}{n \lambda + \mathbb{E}(\sigma^{-2})} \right)^2 \right\}
\end{equation}

@@

\begin{equation}
\label{mf4}
q(\sigma^2) \propto \sigma^{-2(n/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{i=1}^{n}y_i^2 + n\mathbb{E}(\mu^2) - 2\sum_{i=1}^{n} y_i \mathbb{E}(\mu)) - \beta}{\sigma^2} \right\}.
\end{equation}

It is evident that $q(\mu | \lambda) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\lambda})$ and $q(\sigma^2 | \lambda) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$. The parameters $\lambda = (\tilde{\gamma}, \tilde{\lambda}, \tilde{\alpha}, \tilde{\beta})'$ can be found by the set of mean field equations:

\begin{align}
\tilde{\gamma} &= \frac{\alpha / \beta \gamma + \lambda \sum_{i=1}^{n} y_i} {n \lambda + \alpha / \beta} \label{mf5} \\ 
\tilde{\lambda} &= \frac{n \alpha / \beta}{n \lambda + \alpha / \beta} \\
\tilde{\alpha} &= \frac{n}{2} + \alpha \\
\tilde{\beta} &= \frac{1}{2} \left(\sum_{i=1}^{n} y_i^2 + n(\tilde{\gamma}^2 + \tilde{\lambda}) - 2 \sum_{i=1}^{n} y_i \tilde{\gamma} \right) + \beta. \label{mf6}
\end{align}.

@@

In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $\tilde{q_i(\theta_i|\lambda_i)}.$ One method in \citet{Friston2006} uses a Laplace approximation to substitute a Gaussian distribution for these unrecognizable $q_i(\theta_i | \lambda_i)$, though the only requirement is that the substitute has an expectation that is a tractable function of its parameters to be used for other parameters in (\ref{mf2}).

Matching distributions in this way is a very similar method to finding the posterior conditional distribution, $p(\theta_i | y, \theta_{\setminus i}$ in a Gibbs MCMC scheme, with dependence on other parameters replaced by their expectations. The optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, if it exists in a recognisable form such as the exponential family. The secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable distributions.
\vspace{5mm}

The required algorithm is known as a coordinate ascent algorithm, and is described below where $\lambda$ is a $p$ dimensional vector.

@@

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density}
 \KwResult{Mean Field Approximation}
 Use (\ref{mf3}) to match each $q(\theta_i|\lambda_i)$ to a tractable distribution or substitute a further approximation if neccesary.\;
 Derive the set of mean field equations, such as (\ref{mf5}) - (\ref{mf6}).\;
 Initialise $\lambda$ randomly. \;
 \While{Not converged}{
  \For{$i =  1$ \KwTo $k$}{
      Hold $\lambda_{j \neq i}$ fixed\;
      Update $\lambda_i$ using the relevant mean field equation.
     }
 }
 \caption{Coordinate Ascent for MFVB}
  \label{alg:algorithm1}
\end{algorithm}

@@

The algorithm for SVB iteratively takes the  derivative of $\mathcal{L}(q(\theta | \lambda), y)$ in (\ref{ELBO}) with respect to $\lambda$ and the following updating step is applied until the ELBO converges to a maximum.

\begin{equation}
\label{SGA1}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} \mathcal{L}(q(\theta | \lambda^{(m)}), y),
\end{equation}

where $\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ is the vector of partial derivatives of $\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ with respect to each element of $\lambda$. This update requires some initial values $\lambda^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate. If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.

\begin{align}
\sum_{m=1}^{\infty} \rho^{(m)} &=  \infty \\
\sum_{m=1}^{\infty} (\rho^{(m)})^2 &<  \infty.
\end{align}

Whilst a global maximum is desired, the ELBO is a high dimensional problem specific function that makes finding the global maximum extremely difficult and often only a local maximum can be found. One option to alleiviate this problem is to start the algorithm at a range of initial values choose the converged value with the highest maximised ELBO and hence lowest KL divergence to the true posterior

SVB can find only the optimal values for $\lambda$ that are optimal for a distributional family $q$ that is specified by the user. To run SVB many choices of $q$ may be used and one selected on the basis of the highest ELBO. $q(\theta | \lambda)$ is restricted to the family of distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\lambda$ and integration with respect to $\theta$ are interchangable, and SVB can only be applied to models where the log likelihood for $y$ is able to be evaluated. Given these conditions \citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be taken by

\begin{equation}
\label{SGA2}
\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda^{(m)}))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda^{(m)})))
\end{equation}

where $s = 1, \dots, S$ indicates simulations from $q(\theta | \lambda^{(m)})$.

@@

\begin{equation}
\label{SGA3}
G_i^{(m)} = \sum_{j = 1}^{m} \left(\nabla_{\lambda_i}\mathcal{L}(q(\theta | \lambda^{(m)})\right)^2,
\end{equation}

@@

The resulting Stochastic Gradient Ascent algorithm proceeds below with a $p$ dimensional $\lambda$ vector.

>>>>

Electricity prices have been subject to an extensive forecasting literature and are known to be extremely volatile. For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of oberservations less than \$100/mWh, but on 39 occaisions the price spiked greater than \$1,000/mWh, reaching as high as \$9,974/mWh on November 29, 2012. For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period can be more important than the mean value. To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it can average over parameter and model uncertainty in a relatively straightforward manner, which can be difficult in frequentist methods. See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for density forecasting in general.

Bayesian methods are computationally complex as they require integration over the set of unknown parameters, which is often of a high dimension in a model that can capture the strong seasonal patterns in short term electricity load \citep{Taylor2003} that is a major driver for changes in electricity load as well as price. To avoid this problem there are a range of techniques available to obtain the predictive distribution which do not involve slow grid based numerical integration, notably Markov Chain Monte Carlo (MCMC) methods. While MCMC has very strong convergence properties, computation is slow for even moderately complex models, which is problematic as electricity demand in Victoria operates on a five minute interval. The predictive density for the next period's electricity load may not be available before it is observed.

Alternatives to MCMC include Variational Bayes (VB), which aim to replace the unknown true posterior distribution for a vector of unknown parameters $\boldsymbol{\theta}$ with a parametric approximation. Variational Bayes will be explored as an alternative computation strategy to MCMC in the context of short term electricity load forecasting using data from the 2011-2015 period in Victoria. The advantages in computation time for VB will be compared to the loss of statistical accuracy associated with the use of an approximation of the posterior instead of the truth.

@@

Exact Bayesian inference in the context of forecasting uses the predictive density of $Y_{T+1}$ conditioned on the observed series, $y_{t}, t = 1, \dots T$, which is denoted by $y_{1:T}$. The first step requires the posterior distribution of the model parameters $\boldsymbol{\theta}$, $p(\boldsymbol{\theta} | y_{1:T})$, where

\begin{equation}
\label{posterior}
 p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}}{\int_{\boldsymbol{\theta}}p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
\end{equation}

Generally the solution to (\ref{posterior}) is unavailable. This section will offer two alternatives, MCMC and VB.
MCMC is used to create a sample from $p(\boldsymbol{\theta} | y_{1:T})$, and any function of interest from $p(\boldsymbol{\theta} | y_{1:T})$ can be found with the relevant function applied to the sample, while VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxillary parameters associated with the approximation and may depend on the observations $y_{1:T}$. For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian $\boldsymbol{\lambda}$ would be the mean and variance of $\boldsymbol{\theta}$. Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ such that the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ is minimised.

@@

There are many types of MCMC algorithm, one of the most common of these is the Gibbs sampler, which iteratively samples through each element of a $k$ dimensional parameter vector $\boldsymbol{\theta}$ via the conditional distributions

@@

With enough iterations, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$ and can be used for inference. The computation time per iteration and number of iterations required for convergence is problem specific but typically increases with model complexity.

@@

\begin{equation}
\label{AR2}
y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t
\end{equation}

where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. The likelihood of parameters $\boldsymbol{\theta} = (\phi_1, \phi_2, \sigma^2)'$ is given by

\begin{align}
\label{likelihood}
L(\boldsymbol{\theta} | y_{1:T}) &= p(y_1, y_2 | \boldsymbol{\theta}) \prod_{t=3}^{T}p(y_t | y_{1:t-1},\boldsymbol{\theta}) \nonumber \\
&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( \textbf{y}' \Sigma^{-1} \textbf{y} + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}
\end{align}

@@

\begin{equation}
\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right],\
\end{equation}

@@

\begin{align}
p(\sigma^2) &\propto \sigma^{-2} \nonumber \\
p(\phi_1, \phi_2) &\propto \mathbb{I}(\phi_2 > -1)\mathbb{I}(\phi_2 < 1 + \phi_1) \mathbb{I}(\phi_2 < 1 - \phi_1) \nonumber
\end{align}

are used, where $\mathbb{I}$ is the indicator function that equals one when the condition in the brackets is true and zero otherwise. This prior for $p(\phi_1, \phi_2)$ defines the AR(2) stationary region and is independent to $p(\sigma^2)$.

The conditional distribution $p(\sigma^2 | y_{1:T}, \phi_1, \phi_2)$ is Inverse Gamma, while the conditionals for both $\phi$ parameters are intractable and sampled using a Metropolis-Hastings step with truncated normal random walk proposals. The truncation ensures that draws are only taken from the AR(2) stationary region.

@@

A faster alternative to MCMC is Variational Bayes, which introduces an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ choosing the functional form $q$ and given auxillary parameters $\boldsymbol{\lambda}$ so that the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$ is minimised. The KL divergence is defined by

\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})||p(\boldsymbol{\theta} | y_{1:T}] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
\end{equation}

The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that will equal zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}. Note that $KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})||p(\boldsymbol{\theta} | y_{1:T})]$ can be expressed as

\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})||p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q, y_{1:T})
\end{equation}

where $\mathcal{L}(q, y_{1:T})$ is known as the Evidence Lower Bound (ELBO), defined by

\begin{equation}
\label{ELBO}
\mathcal{L}(q, y_{1:T}) = \int_{\boldsymbol{\theta}} q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int_{\boldsymbol{\theta}} q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
\end{equation}

Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}). Maximising the ELBO is much more convenient than minimising the KL Divergence, which can be seen in the following sections based on the two major implementations of Varitaional Bayes: Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these take advantage of the functional form of the ELBO and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.

@@

\begin{equation}
\label{mf1}
q(\theta|\lambda) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
\end{equation}

Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars. Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ which may be of a different dimension to $\theta_i$. $\boldsymbol{\lambda_i}$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\boldsymbol{\lambda_i})$.

MFVB is widely used as it greatly simplifies maximisation of the ELBO, particularly in exponential family models (\citealp{Jordan1999}; \citealp{Ghahramani2000}, \citealp{Wainwright2008}).  Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the computation is simple. MFVB expresses (\ref{ELBO}) as a function of one $q_i$ only and then maximises with respect to that $q_i$. The final algorithm requires this to be repeated for each $i$ in $1, \dots, k$. Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that

\begin{equation}
\label{mf2}
q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y_{1:T},\boldsymbol{\theta}))])
\end{equation}

For illustration, consider data $y_t$ for $t = 1, \dots, T$ generated from a $\mathcal{N}(\mu, \sigma^2)$ distribution with priors $p(\mu) \sim \mathcal{N}(\gamma, \tau)$ and $p(\sigma^2) \sim Inv.Gamma(\alpha, \beta)$. It can be shown that

\begin{equation}
\label{mf3}
q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T\tau + \mathbb{E}_{q(\sigma^2)}(\sigma^{-2}))}{2\tau\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\gamma + \lambda \sum_{t=1}^{T} y_t}{T \tau + \mathbb{E}_{q(\sigma^2 )}(\sigma^{-2})} \right)^2 \right\}
\end{equation}

@@

\begin{equation}
\label{mf4}
q(\sigma^2 | \boldsymbol{\lambda}_2) \propto \sigma^{-2(T/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{t=1}^{T}y_t^2 + n\mathbb{E}_{q(\mu)}(\mu^2) - 2\sum_{t=1}^{T} y_t \mathbb{E}_{q(\mu)}(\mu)) - \beta}{\sigma^2} \right\}.
\end{equation}

It is evident that $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$. The parameters $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$ can be found by the set of mean field equations:

\begin{align}
\tilde{\gamma} &= \frac{\tilde{\alpha} / \tilde{\beta} \gamma + \tau \sum_{t=1}^{T} y_t} {T \lambda + \tilde{\alpha} / \tilde{\beta}} \label{mf5} \\ 
\tilde{\tau} &= \frac{T \tilde{\alpha} / \tilde{\beta}}{T \tau + \tilde{\alpha} / \tilde{\beta}} \label{mf6} \\
\tilde{\alpha} &= \frac{T}{2} + \alpha  \label{mf7} \\
\tilde{\beta} &= \frac{1}{2} \left(\sum_{t=1}^{T} y_t^2 + T(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \sum_{t=1}^{T} y_t \tilde{\gamma} \right) + \beta. \label{mf8}
\end{align}.

@@

In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ One method in \citet{Friston2006} uses a Laplace approximation for these unrecognizable $q_i(\theta_i |\boldsymbol{\lambda_i})$ by matching the mode and inverse hessian matrix to a gaussian distribution, though the only requirement is that the substitute has an expectation that is a tractable function of its parameters to be used for other parameters in (\ref{mf2}).

Matching distributions in this way is a very similar method to finding the posterior conditional distribution, $p(\theta_i | \boldsymbol{\theta}_{j \neq i}, y_{1:T})$ in a Gibbs MCMC scheme, with dependence on other parameters replaced by their expectations. The optimal approximating family for $\theta_i$, $q_i(\theta_i |\boldsymbol{\lambda_i})$, will come from the same distributional family as the conditional distribution $p(\theta_i | \boldsymbol{\theta}_{j \neq i}, y_{1:T})$, if it exists in a recognisable form such as the exponential family. The secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable distributions.
\vspace{5mm}

The required algorithm is known as a coordinate ascent algorithm, and is described below for the model discussed above

@@

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density}
 \KwResult{Mean Field Approximation}
 Use (\ref{mf3}) to match $q(\mu|\boldsymbol{\lambda}_1)$ to a Gaussian distributon and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to an Inverse Gamma distribution\;
 Anaytically derive the set of mean field equations in \ref{mf5}) - (\ref{mf8})\;
 Initialise $\boldsymbol{\lambda}^{(1)}$ randomly.\;
 Evaluate $\mathcal{L}(q, y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
 Set $\tilde{\alpha}$ using \ref{mf7}.\;
 \While{$\mathcal{L}(q, y_{1:T})^{(m)} - \mathcal{L}(q, y_{1:T})^{(m-1)} > \epsilon$}{
  Set $\tilde{\gamma}^{(m)}$ using \ref{mf5}.\;
  Set $\tilde{\tau}^{(m)}$ using \ref{mf6}.\;
  Set $\tilde{\beta}^{(m)}$ using \ref{mf8}.\;
  Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$.\;
  Set $m = m + 1$.\;
 }
 \caption{Coordinate Ascent for MFVB}
  \label{alg:algorithm1}
\end{algorithm}

@@

The algorithm for SVB iteratively takes the  derivative of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ in (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ and the following updating step is applied until the ELBO converges to a maximum.

\begin{equation}
\label{SGA1}
\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}),
\end{equation}

where $\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$. This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate. If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.

\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}

Whilst a global maximum is desired, Variational Bayes theory does not offer any insight to the shape of the ELBO so determining if a local maximum is a global maximum is difficult. One option to alleiviate this problem is to start the algorithm at a range of initial values choose the converged value with the highest maximised ELBO and hence lowest KL divergence to the true posterior.

SVB can find only the optimal values for $\boldsymbol{\lambda}$ for a user specified distributional family $q$, and to run SVB many choices of $q$ may be used and one selected on the basis of the highest ELBO. $q$ is restricted to the family of distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\boldsymbol{\lambda}$ and integration with respect to $\boldsymbol{\theta}$ are interchangable, and SVB can only be applied to models where the log likelihood for $y_{1:T}$ is able to be evaluated. Given these conditions \citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be taken by

\begin{equation}
\label{SGA2}
\nabla_{\lambda}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)}))] (\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)})))
\end{equation}

where $s = 1, \dots, S$ indicates simulations from $q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})$.

@@

\begin{equation}
\label{SGA3}
G_i^{(m)} = \sum_{j = 1}^{m} \left(\nabla_{\lambda_i}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(j)}), y_{1:T})\right)^2,
\end{equation}

@@

The resulting Stochastic Gradient Ascent algorithm proceeds below with a $p$ dimensional $\boldsymbol{\lambda}$ vector.
